\large Neural, Tree, and Symbolic Optimization of Cosmological Particle-Mesh Simulation \vspace{0.3em}

### 2023-07-13 docs/papers/sto/so.tex
FIXUP draft minor
FIXUP draft some updates

\section*{TODO}
\begin{itemize}
\item (paper) Moving some of the following ongoing and completed tasks
  to the main text and add preliminary figures.
\item (data; YL) Incorporate 3lpt branch; testing; find a\_start for
  different mass resolutions; probably replace linear snapshot spacing
  according to the nonlinearity (estimate (with PM till convergence) the
  std acc over the std vel) of each realization.
\item (model; YL) Change nbody for loop by lax scan and/or while loops.
  while\_loop can help to save compilation when trained on different
  snapshots. Performance difference between scan and while?
\item (model; YL) Compare one pair of \GADGET\ and \pmwd\ (with say 100 to 1000 time
    steps) results and make sure that they agree perfectly on the large
    scales. For this purpose the box size should be large enough (in
    linear regime), say 1 Gpc/$h$.\newline
    \textbf{Update}: for Sobol=6, boxsize=983Mpc, \GADGET\ and \pmwd\ are visually
    consistent at the field level. To further check power spectrum. For odd mesh
    shapes (e.g. 1, 3, 5), pmwd and \GADGET\ are perfectly consistent on large
    scales. \textbf{While for even mesh shapes (e.g. 2, 4), the powers look
    slightly different. To figure out this weird issue.}
\item (model; YL) Probably necessary to first incorporate List \& Hahn's
  recent improvement on FastPM integrator arXiv:2301:09655.
\item (train, test) For $\vs(\vq)$, $\vv(\vq)$, $\delta(\vx)$,
  $(1+\delta)\vv(\vx)$, $(1+\delta)\vv\vv(\vx)$, show $T - 1$ in symlog
  with 0.01 linthresh, and $1-r^2$ in log at 3 different $a$'s; Show $f$
  and $g$ (along axis and face diagonal) at 3 different $a$'s; Track
  mean epoch loss
\item (train) Randomize snapshots of each simulation instead of
  sequential; alternatively accumulate their gradients; Normalize loss
  by number of time steps\newline
  \YZ{\textbf{Update}: Normalizing loss by number of time steps seems to halt
  the training of late time snapshots, i.e. they stop improving.}
\item (train, test) 3 sets of grid for ptcl, force mesh, and loss. Add
  random offset in loss grid (see \autoref{tab:param}) to remove the
  position dependence in loss; likewise for force mesh? how about only
  training (for lower loss), only ``inference'' (for higher accuracy),
  or both? the random force mesh offset can also be replaced by a
  structured scheme, e.g., alternating between 2 interlocking grids
\item (train) Explore if we can use a finer loss grid, from 1x to say
  1.5x and even 2x.
\item (train, test) Non-integer multiple mesh shapes: for uniform
  particle grid, test (with double precision on CPUs) whether they start
  moving by artifact forces on cell scales, in 3 cases: no
  deconvolution, with CIC deconvolution, with SO.
\item (train) find the best loss function for production training
\item (train) For generalizability and robustness, consider testing
  weight decay and dropout. Also they're worth adding if they give lower
  losses at convergence. Note that 512 realization are not too many, and
  we have features that can be very correlated.
\item (train) optuna hyperparam search on a subset of training data
  before production training
  (\href{https://github.com/optuna/optuna-examples/blob/main/haiku/haiku_simple.py}{optuna example}).
\item (train) Reduce learning rate on (mean epoch) loss plateau in
  production
\item (test) Interpolation test: compare SO DM and halo fields and
  summary stats with those of raw pmwd, deconvolved pmwd, and \GADGET,
  on the last 9 test cases and noncubic box.
\item (test) Extrapolation test: demonstrate generalization capability,
  compare extrap PySR, extrap SO, raw pmwd, deconv pmwd, and \GADGET.
\item (test) Generate a sample of $(X, y)$ data with the trained neural network
  for feature selection and symbolic regression.
\item Fix cross references and other journal quirks as a patch
\end{itemize}

@@

\vspace{1em}
\section{Results}

\vspace{1em}
\section{Discussion}

@@

\begin{table}
\centering
\caption{Possible features that can affect force sharpening. TODO}
\label{tab:feat}
\begin{tabular}{rl}
\toprule
type & feature \\
\midrule
& $R_\mathrm{d}$: RMS linear displacement \\
\cmidrule(lr){2-2}
& $R_P$: $\Delta_\linear^2(R_P^{-1}) \triangleq 1$ \\
\cmidrule(lr){2-2}
nonlinear scales & $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$ \\
\cmidrule(lr){2-2}
& $R_\gauss$: $\sigma_\gauss^2(R_\tophat) \triangleq 1$ \\
\cmidrule(lr){2-2}
& $\d R / \d\lna$: time derivatives of the above \\
\midrule
& particle spacing \\
\cmidrule(lr){2-2}
other scales & $l$: PM force mesh cell size \\
\cmidrule(lr){2-2}
& softening length \\
\midrule
& $G_m(a) \triangleq D_m(a) / a^m, m \in \{1, 2, \YL{3's?}\}$ \\
\cmidrule(lr){2-2}
& $\d\ln G_m / \d\lna$ \\
\cmidrule(lr){2-2}
dimensionless factors & $\Omegam(a)$ \\
\cmidrule(lr){2-2}
& $\d\ln\!H / \d\lna$ \\
\cmidrule(lr){2-2}
& $\Delta\lna$: time step size in $\lna$ \\
\bottomrule
\end{tabular}
\end{table}

>>>>

\section*{TODO}
\begin{itemize}
\item (paper) Moving some of the following ongoing and completed tasks
  to the main text and add preliminary figures.
\item (data; YL) Incorporate 3lpt branch; testing; find a\_start for
  different mass resolutions; probably replace linear snapshot spacing
  according to the nonlinearity (estimate (with PM till convergence) the
  std acc over the std vel) of each realization.
\item (model; YL) Change nbody for loop by lax scan and/or while loops.
  while\_loop can help to save compilation when trained on different
  snapshots. Performance difference between scan and while?
\item (model; YL) Compare one pair of \GADGET\ and \pmwd\ (with say 100 to 1000 time
    steps) results and make sure that they agree perfectly on the large
    scales. For this purpose the box size should be large enough (in
    linear regime), say 1 Gpc/$h$.\newline
    \YZ{\textbf{Update}: for Sobol=6, boxsize=983Mpc, \GADGET\ and \pmwd\ are
    visually consistent at the field level. For power spectrum: For
    odd mesh shapes (e.g. 1, 3, 5), pmwd and \GADGET\ are perfectly consistent
    on large scales. \textbf{While for even mesh shapes (e.g. 2, 4), the powers
    look slightly different. To figure out this weird issue.}}
\item (model; YL) Probably necessary to first incorporate List \& Hahn's
  recent improvement on FastPM integrator arXiv:2301:09655.
\item (train, test) For $\vs(\vq)$, $\vv(\vq)$, $\delta(\vx)$,
  $(1+\delta)\vv(\vx)$, $(1+\delta)\vv\vv(\vx)$, show $T - 1$ in symlog
  with 0.01 linthresh, and $1-r^2$ in log at 3 different $a$'s; Show $f$
  and $g$ (along axis and face diagonal) at 3 different $a$'s; Track
  mean epoch loss
\item (train) Randomize snapshots of each simulation instead of
  sequential; alternatively accumulate their gradients; Normalize loss
  by number of time steps\newline
  \YZ{\textbf{Update}: Normalizing loss by number of time steps seems to halt
  the training of late time snapshots, i.e. they stop improving.}
\item (train, test) 3 sets of grid for ptcl, force mesh, and loss. Add
  random offset in loss grid (see \autoref{tab:param}) to remove the
  position dependence in loss; likewise for force mesh? how about only
  training (for lower loss), only ``inference'' (for higher accuracy),
  or both? the random force mesh offset can also be replaced by a
  structured scheme, e.g., alternating between 2 interlocking grids
\item (train) Explore if we can use a finer loss grid, from 1x to say
  1.5x and even 2x.
\item (train, test) Non-integer multiple mesh shapes: for uniform
  particle grid, test (with double precision on CPUs) whether they start
  moving by artifact forces on cell scales, in 3 cases: no
  deconvolution, with CIC deconvolution, with SO.
\item (train) find the best loss function for production training
\item (train) For generalizability and robustness, consider testing
  weight decay and dropout. Also they're worth adding if they give lower
  losses at convergence. Note that 512 realization are not too many, and
  we have features that can be very correlated.\newline
  \YZ{Dropout with Flax has been coded. We can use Optax AdamW for weight decay.}
\item (train) optuna hyperparam search on a subset of training data
  before production training
  (\href{https://github.com/optuna/optuna-examples/blob/main/haiku/haiku_simple.py}{optuna
  example}).\newline
  \YZ{Do we also use another subset of the training data as validation data?}
\item (train) Reduce learning rate on (mean epoch) loss plateau in
  production\newline
  \YZ{A simple scheduler has been coded, to be fine-tuned for production training.}
\item (test) Interpolation test: compare SO DM and halo fields and
  summary stats with those of raw pmwd, deconvolved pmwd, and \GADGET,
  on the last 9 test cases and noncubic box.
\item (test) Extrapolation test: demonstrate generalization capability,
  compare extrap PySR, extrap SO, raw pmwd, deconv pmwd, and \GADGET.
\item (test) Generate a sample of $(X, y)$ data with the trained neural network
  for feature selection and symbolic regression.
\item (paper) Think about what figures and tables to show in the paper. Update
to 6 items could be included. Some tables could be included in the context as
list if we have too many figures to show?\newline
  \YZ{figures:
  1. $f$ and $g$ functions, both NN and Symbolic.
  2. Density field (and halo?) with zoom in for small scale sharpening.
  3. Transfer function and correlation coefficient.
  4. Halo related stats.}
\item Fix cross references and other journal quirks as a patch
\end{itemize}

@@

\vspace{1em}
\section{Results}
% Results should be divided by topical subheadings

\subsection{Trained SO neural networks and symbolic expressions}

\subsection{Dark matter fields}

\subsection{Halo catalogs}

\vspace{1em}
\section{Discussion}
% Discussion does not contain subheadings

@@

\begin{table}
\centering
\caption{TODO Possible features that can affect force sharpening.
All scales are then multiplied by the wavenumber $k$ to form dimensionless
factors.}
\label{tab:feat}
\begin{tabular}{rl}
\toprule
type & feature \\
\midrule
& $R_\mathrm{d}$: RMS linear displacement \\
\cmidrule(lr){2-2}
& $R_P$: $\Delta_\linear^2(R_P^{-1}) \triangleq 1$ \\
\cmidrule(lr){2-2}
nonlinear scales & $R_\tophat$: $\sigma_\tophat^2(R_\tophat) \triangleq 1$ \\
\cmidrule(lr){2-2}
& $R_\gauss$: $\sigma_\gauss^2(R_\tophat) \triangleq 1$ \\
\cmidrule(lr){2-2}
& $\d R / \d\lna$: time derivatives of the above \\
\midrule
& particle spacing \\
\cmidrule(lr){2-2}
other scales & $l$: PM force mesh cell size \\
\cmidrule(lr){2-2}
& softening length \\
\midrule
& $G_m(a) \triangleq D_m(a) / a^m, m \in \{1, 2, \YL{3's?}\}$ \\
\cmidrule(lr){2-2}
& $\d\ln G_m / \d\lna$ \\
\cmidrule(lr){2-2}
dimensionless factors & $\Omegam(a)$ \\
\cmidrule(lr){2-2}
& $\d\ln\!H / \d\lna$ \\
\cmidrule(lr){2-2}
& $\Delta\lna$: time step size in $\lna$ \\
\bottomrule
\end{tabular}
\end{table}
